Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-30 submission/manuscript.Rmd
res 3 and 4 edits

Mutations in numerous chromatin-modifying enzymes cause neurodevelopmental disorders (NDDs). Loss of repressive chromatin regulators can lead to the aberrant transcription of tissue-specific genes outside of their intended context, however the mechanisms and consequences of their dysregulation are largely unknown. Here, we explored this cellular identity crisis in NDDs by investigating lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3), in tissue identity. We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key behavioral phenotypes of Claes-Jensen X-linked intellectual disability, aberrantly expresses many liver, muscle, ovary, and testis genes within the amygdala and hippocampus. Gonad-enriched genes expressed in the _Kdm5c_-KO brain were typically unique to germ cells, indicating an erosion of the soma-germline boundary. Germline genes are usually decommissioned in somatic lineages in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_. Germline gene suppression is sexually dimorphic, as female EpiLCs required a higher dose of KDM5C to maintain germline gene suppression. Using a curated list of mouse germline-enriched genes, we found KDM5C is selectively recruited to a subset of germline gene promoters that contain CpG islands (CGIs) to facilitate DNA CpG methylation (CpGme) during ESC to EpiLC differentiation. However, late stage spermatogenesis genes devoid of promoter CGIs can also become activated in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors. Thus, distinct suppressive mechanisms are recruited to different germline gene classes and germline transcriptional programs can mirror germ cell differentiation in the background of typical somatic tissue development.

@@

A single genome holds the instructions to generate the myriad of cell types found an organism. This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through DNA and histone modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001; @lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017]. Human genetic studies revealed mutations in chromatin regulators are a major cause of neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018] and many studies have identified their importance in regulating the expression of brain-specific genes. Loss of some chromatin regulators can also result in the ectopic transcription of tissue-specific genes outside of their target environment, such as the misexpression of liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009]. However, the mechanisms underlying ectopic gene expression and its impact upon neurodevelopment are poorly understood.

To elucidate the role of tissue identity in chromatin-linked NDDs, it is essential to first characterize the nature of the dysregulated genes and the molecular mechanisms governing their de-repression. Here we focus on lysine demethylase 5C (KDM5C), also known as SMCX or JARID1C, which erases histone 3 lysine 4 di- and trimethylation (H3K4me2/3), a permissive chromatin modification enriched at gene promoters[@iwaseXLinkedMentalRetardation2007]. Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534). MRXSCJ is more common and severe in males and its neurological phenotypes include intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020]. Male _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ phenotypes, including hyperaggression, increased seizure propensity, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017]. RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus revealed ectopic expression of some germline genes within the brain[@scandagliaLossKdm5cCauses2017]. However, it is unknown whether other types of testis genes are dysregulated, at what point in development germline gene misexpression begins, and if other tissue-specific genes are aberrantly transcribed with KDM5C loss.

Distinguishing between germ cells and somatic cells is a key feature of multicellularity[@devlinPanmetazoanViewGermlinesoma2023] that occurs during early embryogenesis in many metazoans[@lehmannGermlineStemCells2012]. In mammals, chromatin regulators are crucial for decommissioning germline genes in somatic cells during the transition from na&#239;ve to primed pluripotency. Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells and are then decorated with DNA CpG methylation (CpGme) in the post-implantation embryo[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021]. The contribution of KDM5C to this process remains unclear. It is additionally unknown if KDM5C-mediated germline gene repression is impacted by chromosomal sex, as females have a higher KDM5C dose than males due to its escape from X chromosome inactivation[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008]. Furthermore, studies on germline gene repression have primarily focused on marker genes important for germ cell development rather than germline genes as a whole, given the lack of a curated list for germline-enriched genes. Therefore, it is unknown if the mechanism of repression differs for certain classes of germline genes, e.g. meiotic genes versus spermatid differentiation genes.

To illuminate KDM5C's role in tissue identity, here we characterized the aberrant expression of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like stem cells (EpiLCs), an _in vitro_ model of the post-implantation embryo. We curated list of mouse germline-enriched genes, which enabled genome-wide analysis of germline gene silencing mechanisms for the first time. Based on the data presented below, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.

@@

Previous RNA sequencing (RNA-seq) of the adult male _Kdm5c_-KO hippocampus revealed ectopic expression of some germline genes unique to the testis[@scandagliaLossKdm5cCauses2017]. It is currently unclear if the testis is the only tissue type misexpressed in the _Kdm5c_-KO brain. We thus characterized the role of KDM5C in brain tissue identity by systematically assessing the dysregulation of genes specific to or enriched in 17 mouse tissues[@liComprehensiveMouseTranscriptomic2017]. We analyzed our published mRNA-seq datasets of the adult amygdala and hippocampus from wild-type and constitutive _Kdm5c_-KO male mice[@vallianatosMutuallySuppressiveRoles2020].

@@

We found a large proportion significantly upregulated genes within the _Kdm5c_-KO brain (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) are typically enriched in non-brain tissues (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B). For both the amygdala and hippocampus, the majority of tissue-enriched differentially expressed genes (DEGs) were testis genes (Figure 1A-C). Even though the testis has the largest total number of tissue-biased genes (`r gettissval("count", "Testis")` genes) compared to any other tissue, testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Testis"), 3)`, Fisher's Exact Test). One example of a testis-enriched gene misexpressed in the _Kdm5c_-KO brain is _FK506 binding protein 6 (Fkbp6)_, a known regulator of PIWI-interacting RNAs (piRNAs) and meiosis[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012] (Figure 1C).

Interestingly, we also observed significant enrichment of ovary-biased DEGs in both the amygdala and hippocampus (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Ovary"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Ovary"), 3)`, Fisher's Exact) (Figure 1A-B). Ovary-enriched DEGs included _Zygotic arrest 1 (Zar1)_, which sequesters mRNAs in oocytes for meiotic maturation[@chengMammalianOocytesStore2022] (Figure 1D). Given that the _Kdm5c_-KO mice we analyzed are male, these data demonstrate that the ectopic expression of tissue-enriched genes is independent of organismal sex.

Although not consistent across brain regions, we also found significant enrichment of DEGs biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 2)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Liver"), 3)`, Fisher's Exact Test) and the muscle (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 2)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Muscle"), 3)`, Fisher's Exact Test) (Figure 1A-B). Among liver-biased DEG derepressed in both the hippocampus and amygdala is _Apolipoprotein C-I (Apoc1)_, a lipoprotein metabolism and transport gene[@roulandRoleApolipoproteinC12022] that has been implicated in Alzheimer's disease[@leducAPOECholesterolHomeostasis2010] (Figure 1E).

For all _Kdm5c_-KO tissue-enriched DEGs, aberrantly expressed mRNAs are polyadenylated and spliced into mature transcripts (Figure 1C-E). Of note, we observed little to no dysregulation of brain-enriched genes (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Brain"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Brain"), 3)`, Fisher's Exact), despite the fact these are brain samples and the brain has the second highest total number of tissue-enriched genes (`r gettissval("count", "Brain")` genes). Altogether, these results suggest the aberrant expression of tissue-enriched genes within the brain is a major effect of KDM5C loss.

@@

In addition to germ cells, the testis also contains both somatic cells, such as Leydig cells that support hormone production and germline functions. To determine if _Kdm5c_-KO results in ectopic expression of somatic testicular genes, we first evaluated their known functions through gene onotology analysis. We found _Kdm5c_-KO testis-enriched DEGs had high enrichment of germline-relevant ontologies, including spermatid development (GO: 0007286, p.adjust = `r signif(getGO("spermatid development", testisGO, 'p.adjust', "no"), 3)`) and sperm axoneme assembly (GO: 0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", testisGO, 'p.adjust', "no"), 3)`) (Figure 2A).

To validate the cell-type origin of testicular DEGs, we then compared their expression in somatic and germ cells within the testis. We first evlauted their expression in wild-type testes versus testes with germ cell depletion[@muellerIndependentSpecializationHuman2013], which was accomplished by heterozygous _W_ and _Wv_ mutations in the enzymatic domain of _c-Kit_ (Kit^W/Wv^)[@handelSertoliCellDifferentiation1979]. Almost all _Kdm5c_-KO testis-enriched DEGs lost expression with germ cell depletion (Figure 2B). We then assessed testis-enriched DEG expression in a published single cell RNA-seq dataset that identified cell type-specific markers within the testis[@greenComprehensiveRoadmapMurine2018]. Some _Kdm5c_-KO testis-enriched DEGs were classified as specific markers for different germ cell developmental stages (e.g. spermatogonia, spermatocytes, round spermatids, and elongating spermatids), yet none marked somatic cells (Figure 2C). Together, these data demonstrate that the _Kdm5c_-KO brain aberrantly expresses germline genes, but not somatic testicular genes, reflecting an erosion of the soma-germline boundary.

As of yet, research on germline silencing mechanisms has focused on a handful of key genes rather than assessing germline gene suppression genome-wide due to the lack of a comprehensive mouse germline gene list. We therefore generated a list of germline-enriched genes using RNA-seq datasets of Kit^W/Wv^ mice that included males and females at embryonic day 12, 14, and 16[@sohGeneRegulatoryProgram2015] and adult male testes[@muellerIndependentSpecializationHuman2013]. We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type gonads 2) their expression in any non-gonadal tissue of adult wild type mice[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of their maximum expression in the wild-type germline, and 3) their expression in the germ cell-depleted gonads, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline. These criteria yielded `r nrow(germgenes)` germline-enriched genes (Figure 2D), which was hereafter used as a resource to globally characterize germline gene misexpression with _Kdm5c_ loss (Supplementary table 1).

@@

@@

Germ cells are typically distinguished from somatic cells soon after the embryo implants into the uterine wall[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014]. A subset of epiblast stem cells become the primordial germ cells (PGCs) while the remainder differentiate into the ectoderm, mesoderm, and endoderm to form the somatic tissues[@bardotMouseGastrulationCoordination2020]. This developmental time point can be modeled _in vitro_ through differentiation of embryonic stem cells (ESCs) into epiblast-like stem cells (EpiLCs) (Figure 3A)[@hayashiReconstitutionMouseGerm2011; @samantaGeneratingPrimedPluripotent2020]. While some germline-enriched genes are also expressed in na&#239;ve embryonic stem cells (ESCs) and in the 2-cell stage[@wellingDAZLRegulatesTet12015; @macfarlanEmbryonicStemCell2012; @suzukiLossMAXResults2016], they are silenced as they differentiate into EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021]. Therefore, we tested if KDM5C was necessary for the initial silencing germline genes in somatic lineages by evaluating the impact of _Kdm5c_ loss in male EpiLCs.

_Kdm5c_-KO cell morpholgy during ESC to EpiLC differentiation appeared normal (Figure 3A) and EpiLCs properly expressed markers of primed pluripotency, such as _Dnmt3b_, _Fgf5_, _Pou3f1_, and _Otx2_ (Figure 3B). We then identified tissue-enriched DEGs in our previously published RNA-seq dataset of wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] (DESeq2, log2 fold change > `r l2fcco`, q < 0.1). Similar to the _Kdm5c_-KO brain, we observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3B).

Using the list of mouse germline-enriched genes curated above, we found `r nrow(epigerm)` germline genes were misexpressed in male _Kdm5c_-KO EpiLCs. We then compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain to determine if germline genes are constitutively dysregulated or change over the course of development. We found the majority of germline DEGs were unique to either EpiLCs or the brain, with only _Cyct_ shared across all tissue/cell types (Figure 3E-F). EpiLCs had particularly high enrichment of meiosis-related gene ontologies (Figure 3G), such as meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO:0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`). While there was modest enrichment of meiotic gene ontologies in both brain regions, the _Kdm5c_-KO hippocampus primarily expressed late-stage spermatogenesis genes involved in sperm axoneme assembly (GO:0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`) and sperm motility (GO:0097722, p.adjust = `r signif(getGO("sperm motility", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`).

Notably, DEGs unique to _Kdm5c_-KO EpiLCs included key drivers of germline identity, such as _Stimulated by retinoic acid 8 (Stra8)_ and _Deleted in azoospermia like (Dazl)_ (Figure 3H). These genes are typically expressed when PGCs commit to the germline fate, but are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019]. Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo[@macfarlanEmbryonicStemCell2012; @guptaGenomewideScreenReveals2023]. However, we did not see derepression of two-cell stage-specific genes, like _Duxf3 (Dux)_ (q = `r signif(duxf3, 3)`) and _Zscan4d_ (q = `r signif(zscan4d, 3)`) (Figure 3H), indicating _Kdm5c_-KO EpiLCs do not revert back to a 2-cell-like state. Altogether, these results demonstrate _Kdm5c_-KO EpiLCs express key drivers of germline identity and meiosis while the _Kdm5c_-KO brain primarily expresses spermiogenesis genes.

>>>>

Mutations in numerous chromatin-modifying enzymes cause neurodevelopmental disorders (NDDs). Loss of repressive chromatin regulators can lead to the aberrant transcription of tissue-specific genes outside of their intended context, however the mechanisms and consequences of their dysregulation are largely unknown. Here, we explored this cellular identity crisis in NDDs by investigating lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3), in tissue identity. We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key behavioral phenotypes of Claes-Jensen X-linked intellectual disability, aberrantly expresses many liver, muscle, ovary, and testis genes within the amygdala and hippocampus. Gonad-enriched genes expressed in the _Kdm5c_-KO brain were typically unique to germ cells, indicating an erosion of the soma-germline boundary. Germline genes are usually decommissioned in somatic lineages in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_. Germline gene suppression is sexually dimorphic, as female EpiLCs required a higher dose of KDM5C to maintain germline gene suppression. Using a curated list of mouse germline-enriched genes, we found KDM5C is selectively recruited to a subset of germline gene promoters that contain CpG islands (CGIs) to facilitate DNA CpG methylation (CpGme) during ESC to EpiLC differentiation. However, late stage spermatogenesis genes devoid of promoter CGIs can also become activated in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors. Thus, distinct suppressive mechanisms are recruited to different germline gene classes and ectopic germline transcriptional programs can mirror germ cell development within somatic tissues.

@@

A single genome holds the instructions to generate the myriad of cell types found within an organism. This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through DNA and histone modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001; @lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017]. Human genetic studies revealed mutations in chromatin regulators are a major cause of neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018] and many studies have identified their importance for regulating brain-specific transcriptional programs. Loss of some chromatin regulators can also result in the ectopic expression of tissue-specific genes outside of their target environment, such as the misexpression of liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009]. However, the mechanisms underlying ectopic gene expression and its impact upon neurodevelopment are poorly understood.

To elucidate the role of tissue identity in chromatin-linked NDDs, it is essential to first characterize the nature of the dysregulated genes and the molecular mechanisms governing their de-repression. Here we focus on lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C), which erases histone 3 lysine 4 di- and trimethylation (H3K4me2/3), a permissive chromatin modification enriched at gene promoters[@iwaseXLinkedMentalRetardation2007]. Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534). MRXSCJ is more common and severe in males and its neurological phenotypes include intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020]. Male _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ phenotypes, including hyperaggression, increased seizure propensity, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017]. RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus surprisingly revealed ectopic expression of some germline genes within the brain[@scandagliaLossKdm5cCauses2017]. However, it is unclear if other tissue-specific genes are aberrantly transcribed with KDM5C loss, at what point in development germline gene misexpression begins, and what mechanisms underlie their dysregulation.

Distinguishing between germ cells and somatic cells is a key feature of multicellularity[@devlinPanmetazoanViewGermlinesoma2023] that occurs during early embryogenesis in many metazoans[@lehmannGermlineStemCells2012]. In mammals, chromatin regulators are crucial for decommissioning germline genes during the transition from na&#239;ve to primed pluripotency. Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells (ESCs) and are then decorated with DNA CpG methylation (CpGme) in the post-implantation embryo[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021]. The contribution of KDM5C to this process remains unclear. Furthermore, studies on germline gene repression have primarily been conducted in males and focused on marker genes important for germ cell development rather than germline genes as a whole, given the lack of a curated list for germline-enriched genes. Therefore, it is unknown if the mechanism of repression differs between sexes or for certain classes of germline genes, e.g. meiotic genes versus spermatid differentiation genes.

To illuminate KDM5C's role in tissue identity, here we characterized the aberrant expression of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like stem cells (EpiLCs), an _in vitro_ model of the post-implantation embryo. We curated list of mouse germline-enriched genes, which enabled genome-wide analysis of germline gene silencing mechanisms for the first time. Based on the data presented below, we propose KDM5C plays a fundamental, sexually dimorphic role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.

@@

Previous RNA sequencing (RNA-seq) of the adult male _Kdm5c_-KO hippocampus revealed ectopic expression of some germline genes unique to the testis[@scandagliaLossKdm5cCauses2017]. It is currently unknown if the testis is the only tissue type misexpressed in the _Kdm5c_-KO brain. We thus characterized the role of KDM5C in brain tissue identity by systematically assessing the dysregulation of genes enriched in 17 mouse tissues[@liComprehensiveMouseTranscriptomic2017]. We analyzed tissue-specific differentially expressed genes (DEGs) in our published mRNA-seq datasets[@vallianatosMutuallySuppressiveRoles2020] of the adult amygdala and hippocampus from wild-type and constitutive _Kdm5c_-KO male mice (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1).

@@

We found a large proportion of significantly upregulated genes within the _Kdm5c_-KO brain are typically enriched within non-brain tissues (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B). For both the amygdala and hippocampus, the majority of tissue-enriched (DEGs) were testis genes (Figure 1A-C). Even though the testis has the largest total number of tissue-biased genes (`r gettissval("count", "Testis")` genes) compared to any other tissue, testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Testis"), 3)`, Fisher's Exact Test). One example of a testis-enriched gene misexpressed in the _Kdm5c_-KO brain is _FK506 binding protein 6 (Fkbp6)_, a known regulator of PIWI-interacting RNAs (piRNAs) and meiosis[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012] (Figure 1C).

Interestingly, we also observed significant enrichment of ovary-biased DEGs in both the amygdala and hippocampus (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Ovary"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Ovary"), 3)`, Fisher's Exact) (Figure 1A-B). Ovary-enriched DEGs included _Zygotic arrest 1 (Zar1)_, which sequesters mRNAs in oocytes for meiotic maturation[@chengMammalianOocytesStore2022] (Figure 1D). Given that the _Kdm5c_-KO mice we analyzed are male, these data demonstrate that the ectopic expression of gonad-enriched genes is independent of organismal sex.

Although not consistent across brain regions, we also found significant enrichment of DEGs biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 2)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Liver"), 3)`, Fisher's Exact Test) and the muscle (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 2)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Muscle"), 3)`, Fisher's Exact Test) (Figure 1A-B). _Apolipoprotein C-I (Apoc1)_ a lipoprotein metabolism and transport gene, is among the liver-biased DEG derepressed in both the hippocampus and amygdala[@roulandRoleApolipoproteinC12022] and its brain overexpression has been implicated in Alzheimer's disease[@leducAPOECholesterolHomeostasis2010] (Figure 1E).

For all _Kdm5c_-KO tissue-enriched DEGs, aberrantly expressed mRNAs are polyadenylated and spliced into mature transcripts (Figure 1C-E). Of note, we observed little to no dysregulation of brain-enriched genes (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Brain"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Brain"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Brain"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Brain"), 3)`, Fisher's Exact), despite the fact these are brain samples and the brain has the second highest total number of tissue-enriched genes (`r gettissval("count", "Brain")` genes). Altogether, these results suggest the aberrant expression of tissue-enriched genes within the brain is a major effect of KDM5C loss.

@@

_Kdm5c_-KO brain expresses testicular germline genes[@scandagliaLossKdm5cCauses2017], however the testis also contains somatic cells that support hormone production and germline functions. To determine if _Kdm5c_-KO results in ectopic expression of somatic testicular genes, we first evaluated the known functions of testicular DEGs through gene onotology. We found _Kdm5c_-KO testis-enriched DEGs had high enrichment of germline-relevant ontologies, including spermatid development (GO: 0007286, p.adjust = `r signif(getGO("spermatid development", testisGO, 'p.adjust', "no"), 3)`) and sperm axoneme assembly (GO: 0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", testisGO, 'p.adjust', "no"), 3)`) (Figure 2A).

We then evlauted testicular DEG expression in wild-type testes versus testes with germ cell depletion[@muellerIndependentSpecializationHuman2013], which was accomplished by heterozygous _W_ and _Wv_ mutations in the enzymatic domain of _c-Kit_ (Kit^W/Wv^)[@handelSertoliCellDifferentiation1979]. Almost all _Kdm5c_-KO testis-enriched DEGs lost expression with germ cell depletion (Figure 2B). We then assessed testis-enriched DEG expression in a published single cell RNA-seq dataset that identified cell type-specific markers within the testis[@greenComprehensiveRoadmapMurine2018]. Some _Kdm5c_-KO testis-enriched DEGs were classified as specific markers for different germ cell developmental stages (e.g. spermatogonia, spermatocytes, round spermatids, and elongating spermatids), yet none marked somatic cells (Figure 2C). Together, these data demonstrate that the _Kdm5c_-KO brain aberrantly expresses germline genes, but not somatic testicular genes, reflecting an erosion of the soma-germline boundary.

As of yet, research on germline gene silencing mechanisms has focused on a handful of key genes rather than assessing germline gene suppression genome-wide due to the lack of a comprehensive gene list. We therefore generated a list of mouse germline-enriched genes using RNA-seq datasets of Kit^W/Wv^ mice that included males and females at embryonic day 12, 14, and 16[@sohGeneRegulatoryProgram2015] and adult male testes[@muellerIndependentSpecializationHuman2013]. We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type gonads 2) their expression in any non-gonadal tissue of adult wild type mice[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of their maximum expression in the wild-type germline, and 3) their expression in the germ cell-depleted gonads, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline. These criteria yielded `r nrow(germgenes)` germline-enriched genes (Figure 2D), which was hereafter used as a resource to globally characterize germline gene misexpression with _Kdm5c_ loss (Supplementary table 1).

@@

getresval <- function(df, col, symbol){
	val <- subset(df, SYMBOL == symbol)
	val <- val[,col]
	return(val)
}

@@

Germ cells are typically distinguished from somatic cells soon after the embryo implants into the uterine wall[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014], when they are silenced in epiblast stem cells that will  differentiate into the ectoderm, mesoderm, and endoderm to form the somatic tissues[@bardotMouseGastrulationCoordination2020]. This developmental time point can be modeled _in vitro_ through differentiation of na&#239;ve embryonic stem cells (nESCs) into epiblast-like stem cells (EpiLCs) (Figure 3A)[@hayashiReconstitutionMouseGerm2011; @samantaGeneratingPrimedPluripotent2020]. While some germline-enriched genes are also expressed in nESCs and in the 2-cell stage[@wellingDAZLRegulatesTet12015; @macfarlanEmbryonicStemCell2012; @suzukiLossMAXResults2016], they are silenced as they differentiate into EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021]. Therefore, we tested if KDM5C was necessary for the initial silencing germline genes in somatic lineages by evaluating the impact of _Kdm5c_ loss in male EpiLCs.

_Kdm5c_-KO cell morpholgy during ESC to EpiLC differentiation appeared normal (Figure 3B) and EpiLCs properly expressed markers of primed pluripotency, such as _Dnmt3b_, _Fgf5_, _Pou3f1_, and _Otx2_ (Figure 3C). We then identified tissue-enriched DEGs in our previously published RNA-seq dataset of wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] (DESeq2, log2 fold change > `r l2fcco`, q < 0.1). Similar to the _Kdm5c_-KO brain, we observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3D). Using our list of mouse germline-enriched genes assembled above, we found `r nrow(epigerm)` germline genes were misexpressed in male _Kdm5c_-KO EpiLCs.

We then compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain to determine if germline genes are constitutively dysregulated or change over the course of development. The majority of germline DEGs were unique to either EpiLCs or the brain, with only _Cyct_ shared across all tissue/cell types (Figure 3E-F). EpiLCs had particularly high enrichment of meiosis-related gene ontologies (Figure 3G), such as meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO:0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`). While there was modest enrichment of meiotic gene ontologies in both brain regions, the _Kdm5c_-KO hippocampus primarily expressed late-stage spermatogenesis genes involved in sperm axoneme assembly (GO:0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`) and sperm motility (GO:0097722, p.adjust = `r signif(getGO("sperm motility", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`).

Notably, DEGs unique to _Kdm5c_-KO EpiLCs included key drivers of germline identity, such as _Stimulated by retinoic acid 8_ (_Stra8_: log2 fold change = `r signif(getresval(epigerm, 'log2FoldChange', 'Stra8'), 3)`, q = `r signif(getresval(epigerm, 'padj', 'Stra8'), 3)`) and _Deleted in azoospermia like_ (_Dazl_: log2 fold change = `r signif(getresval(epigerm, 'log2FoldChange', 'Dazl'), 3)`, q = `r signif(getresval(epigerm, 'padj', 'Dazl'), 3)`) (Figure 3H). These genes are typically expressed when primordial germ cells (PGCs) are committed to the germline fate, but are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019]. Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo[@macfarlanEmbryonicStemCell2012; @guptaGenomewideScreenReveals2023]. However, we did not see derepression of two-cell stage-specific genes, like _Duxf3 (Dux)_ (q = `r signif(duxf3, 3)`) and _Zscan4d_ (q = `r signif(zscan4d, 3)`) (Figure 3H), indicating _Kdm5c_-KO EpiLCs do not revert back to a 2-cell-like state. Altogether, _Kdm5c_-KO EpiLCs expressing key drivers of germline identity and meiosis while the brain primarily expresses spermiogenesis genes indicate germline gene misexpression parallels germline development during proper _Kdm5c_-KO differentiation.
